---
title: "p8105_hw5_ao2965"
author: "Alissa Shams Orchi"
date: "2025-11-14"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)

theme_set(theme_minimal())
```

# Problem 2

### Running the simulations
```{r run-simulations, cache=TRUE}
n = 30
sigma = 5
alpha = 0.05
n_sim = 5000

sim_t_test = function(mu, n = 30, sigma = 5) {
  sim_data = rnorm(n, mean = mu, sd = sigma)
  test = t.test(sim_data, mu = 0)
  broom::tidy(test) |> 
    select(estimate, p.value)
}

sim_mu_0 = map_dfr(1:n_sim, ~ sim_t_test(mu = 0)) |> 
  mutate(mu_true = 0)

sim_mu_1 = map_dfr(1:n_sim, ~ sim_t_test(mu = 1)) |> 
  mutate(mu_true = 1)

sim_mu_2 = map_dfr(1:n_sim, ~ sim_t_test(mu = 2)) |> 
  mutate(mu_true = 2)

sim_mu_3 = map_dfr(1:n_sim, ~ sim_t_test(mu = 3)) |> 
  mutate(mu_true = 3)

sim_mu_4 = map_dfr(1:n_sim, ~ sim_t_test(mu = 4)) |> 
  mutate(mu_true = 4)

sim_mu_5 = map_dfr(1:n_sim, ~ sim_t_test(mu = 5)) |> 
  mutate(mu_true = 5)

sim_mu_6 = map_dfr(1:n_sim, ~ sim_t_test(mu = 6)) |> 
  mutate(mu_true = 6)

sim_results = bind_rows(
  sim_mu_0,
  sim_mu_1,
  sim_mu_2,
  sim_mu_3,
  sim_mu_4,
  sim_mu_5,
  sim_mu_6
)

```

### Generating the power summary and plot
```{r }
power_summary = sim_results |> 
  group_by(mu_true) |> 
  summarize(
    power = mean(p.value < 0.05),
    avg_estimate_all = mean(estimate),
    avg_estimate_rejected = mean(estimate[p.value < 0.05]),
    n_simulations = n()
  )

power_summary
```
```{r}
power_plot = power_summary |> 
  ggplot(aes(x = mu_true, y = power)) +
  geom_point(size = 3) +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Power vs Effect Size",
    x = "True value of μ",
    y = "Power (Proportion of times null was rejected)"
  ) +
  theme_minimal()

power_plot
```

* As effect size increases, power increases in the plot. This indicates a positive association between them. There is a steep increase in power when the true value of μ increases from 1 to 4. After reaching 4, the association plateaus

### Second plot
`````{r estimates-plot-faceted}
estimates_plot = power_summary |> 
  select(mu_true, avg_estimate_all, avg_estimate_rejected) |> 
  pivot_longer(
    cols = starts_with("avg"),
    names_to = "sample_type",
    values_to = "average_estimate"
  ) |> 
  mutate(
    sample_type = case_when(
      sample_type == "avg_estimate_all" ~ "All samples",
      sample_type == "avg_estimate_rejected" ~ "Rejected null only"
    )
  ) |> 
  ggplot(aes(x = mu_true, y = average_estimate)) +
  geom_point(size = 3, color = "blue") +
  geom_line(color = "blue") +
  facet_wrap(~ sample_type, ncol = 2) +  
  labs(
    title = "Average Estimate of μ̂ by True μ",
    x = "True value of μ",
    y = "Average estimate of μ̂"
  ) +
  geom_abline(slope = 1, intercept = 0,   
            linetype = "dashed", alpha = 0.5) +
  theme_minimal()

estimates_plot
`````


* The sample average of μ̂ across tests where the null was rejected is not approximately equal to the true μ, especially for small effect sizes.  This occurs because when the true effect is small, we only reject the null when random variation produces unusually large estimates. This selection bias disappears as power increases because nearly all samples lead to rejection.
 
# Problem 3

### Raw data
```{r, message=FALSE}
homicide_data = read_csv("homicide-data.csv")

n_obs = nrow(homicide_data)
n_vars = ncol(homicide_data)
var_names = names(homicide_data) |> str_c(collapse = ", ")
```

* The Washington Post homicide dataset contains `r format(n_obs, big.mark = ",")` observations 
and `r n_vars` variables. The variables included are: `r var_names`.

### Create city_state variable and summarize

```{r}
homicide_summary = homicide_data |> 
  mutate(
    city_state = str_c(city, ", ", state),
    city_state = str_replace(city_state, "Milwaukee, wI", "Milwaukee, WI"),
    resolution = case_when(
      disposition == "Closed without arrest" ~ "unsolved",
      disposition == "Open/No arrest" ~ "unsolved", 
      disposition == "Closed by arrest" ~ "solved"
    )
  ) |> 
  filter(city_state != "Tulsa, AL") |> 
  group_by(city_state) |> 
  summarize(
    total_homicides = n(),
    unsolved_homicides = sum(resolution == "unsolved"),
    .groups = "drop"
  )

homicide_summary |> 
  arrange(desc(total_homicides))

homicide_summary |> 
  arrange(desc(unsolved_homicides))
```
* After summarizing the total number of homicides, we see that the 5 cities with the highest number of total homicides are Chicago, Philadelphia, Houston, Baltimore, Detroit.

* After summarizing the total number of unsolved_homicides, we see that the 5 cities with the highest number of total unsolved_homicides are Chicago, Baltimore, Houston, Detroit, Philadelphia.

### Baltimore prop.test
`````{r}
run_prop_test = function(unsolved, total) {
  test = prop.test(x = unsolved, n = total)
  broom::tidy(test) |> 
    select(estimate, conf.low, conf.high)
}
baltimore_data = homicide_summary |> 
  filter(city_state == "Baltimore, MD")

baltimore_results = run_prop_test(
  unsolved = baltimore_data |> pull(unsolved_homicides),
  total = baltimore_data |> pull(total_homicides)
)
baltimore_results
`````

### All cities prop.test

```{r}
all_cities = homicide_summary |> 
  mutate(
    test_results = map2(
      unsolved_homicides, 
      total_homicides, 
      run_prop_test
    )
  ) |> 
  unnest(test_results)

all_cities
```
### Create Plot
```{r, fig.height=10, fig.width=8}
homicide_plot = all_cities |> 
  mutate(city_state = fct_reorder(city_state, estimate)) |> 
  ggplot(aes(x = city_state, y = estimate)) +
  geom_point(size = 2, color = "darkblue") +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.3,
    alpha = 0.7
  ) +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    title = "Proportion of Unsolved Homicides by City",
    subtitle = "Error bars represent 95% confidence intervals",
    x = "",
    y = "Proportion of Unsolved Homicides"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 7),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  )

homicide_plot
```
